---
title: "Birth to 3 County Report"
#author: "Child Poverty Action Lab"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
#description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
#github-repo: childpovertyactionlab/tractmap
favicon: "images/favicon.ico"
---

# Dallas County Census Tract Map {-}

The map below contains a map coloring all census tracts in Dallas county based on the demographic group that is the majority of the population in the respective tract.
The majority group is generated by selecting the population which makes up at least 45% of the total population in the census tract.
The group __MULTI__ is selected whenever there is no demographic group which makes up at least 45% of the population in a given census tract.

```{r, include=FALSE}
rm(list=ls(all=TRUE))
#### Libraries of interest to load ######################
library(tidyverse)
library(tidycensus)
library(lubridate)
library(rio)
library(sf)
library(ggthemes)
library(leaflet)
library(htmltools)

#### List of CPAL Colors to use ########################
CPAL.colors = c("#008097", "#ec008c", "#eaca2d", "#b4b4b4", "#9bd9e3", "#fdddd7")

#### CPAL ggplot theme #################################
theme_cpal <- function(base_size = 12, base_family = "sans") {
  colors <- deframe(ggthemes::ggthemes_data[["fivethirtyeight"]])
  (theme_foundation(base_size = base_size, base_family = base_family)
    + theme(
      line = element_line(colour = "#b4b4b4"),
      rect = element_rect(fill = "#ffffff",
                          linetype = 1, colour = NA),
      text = element_text(family = "Roboto", face = "bold", colour = "#6c6c6c"),
      axis.title = element_text(),
      axis.title.x = element_text(vjust = 2),
      axis.title.y = element_text(vjust = 2),
      axis.text = element_text(color = "#b4b4b4"),
      axis.ticks = element_blank(),
      #axis.ticks.length = unit(6, "pt"),
      axis.line = element_line(color = "#b4b4b4", size = 1.5, linetype = "solid"),
      legend.background = element_rect(),
      legend.position = "none",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      panel.grid.major = element_line(colour = "#e1e1e1"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(hjust = 0, size = rel(1.5), face = "bold"),
      plot.margin = unit(c(1, 2, 1, 1), "lines"),
      panel.border = element_rect(size=1, fill = NA),
      strip.background = element_rect()
    ))
}

#### TidyCensus pull code chunks #######################
acs_b <- load_variables(2019, "acs5")
acs_s <- load_variables(2019, "acs5/subject")
#fips <- tidycensus::fips_codes
counties <- c("Dallas County", 
              "Rockwall County", 
              "Collin County", 
              "Denton County", 
              "Tarrant County", 
              "Kaufman County", 
              "Ellis County", 
              "Johnson County")

acs_var <- c(
  tot_pop = "B01003_001", #total population
  his_pop = "B03002_012", #hispanic population
  wh_pop = "B03002_003", #white population
  bl_pop = "B03002_004", #black population
  as_pop = "B03002_006", #asian population
  pop_3t4 = "B09001_004", #population between 3 to 4
  med_inc = "B19013_001", #median household income
  gini = "B19083_001", #gini coefficient
  pubas_tot = "B22002_002", #total population for table 09010
  pubas_u18 = "B22002_003", #receiving assistance with children under 18 in households
  pop_deg = "B23006_001", #population for highest education variables
  hs_deg = "B23006_009", #total with high school degree
  ba_deg = "B23006_023", #total with bachelors degree or higher
  avg_hhs = "B25010_001", #average household size
  avg_rhhs= "B25010_003", #average renter-occupied household size
  avg_ohhs = "B25010_002", #average owner-occupied household size
  oohh = "B25106_002", #owner-occupied households
  rohh = "B25106_024", #renter occupied households
  thh = "B25106_001", #total households
  es_3t4 = "B14003_004", #enrolled in school 3 to 4
  pop_bp = "S1701_C02_001", #population below poverty
  bp_u18 = "S1701_C02_002", #population under 18 below poverty
  pop_u18 = "S0101_C01_022", #population under 18
  gran_u6 = "B10001_002", #grandchildren under 6 living in grandparent household,
  pop_u5 = "S1701_C01_002", #total population under 5
  bp_u5 = "S1701_C02_003", #total population below poverty under 5
  wra = "B13010_001", #total women between 15 and 50
  births = "B13010_002", #women who've had a birth in the last 12 months 15 to 50
  birmar_u100 = "B13010_004", #women 15 to 50 married under 100% FPL given birth in the last year
  birunm_u100 = "B13010_008" #women 15 to 50 unmarried under 100% FPL given birth in the last year
)

tidypull <- get_acs(
  geography = "tract",
  state = "TX",
  county = "Dallas County",
  variables = acs_var,
  year = 2019, 
  survey = "acs5", 
  output = "wide",
  geometry = TRUE)

dallascounty <- tigris::counties(state = "TX") %>%
  filter(COUNTYFP == 113)

#### Import Childcare Desert Data ################
sub_ccc <- st_read("E:/CPAL Dropbox/Data Library/Texas Department of Family and Protective Services/ChildcareCenters_SubsidizedTX_Feb2021.shp") %>%
  st_transform(crs = 4269)
sta_ccc <- st_read("E:/CPAL Dropbox/Data Library/Texas Department of Family and Protective Services/ChildcareCenters_UnSubsidizedTX_Feb2021.shp") %>%
  st_transform(crs = 4269)

tidyadd <- tidypull %>%
  mutate(sta_ccc = lengths(st_intersects(., sta_ccc)),
         sub_ccc = lengths(st_intersects(., sub_ccc)),
         tot_ccc = sta_ccc+sub_ccc)

#### Cut and Jenks functions ########
jenkscut <- function(x, n) {
  cut(x, BAMMtools::getJenksBreaks(x, k = n), labels = seq_len(n),
      include.lowest = TRUE)
}

sdcut <- function(x) {
  sd1p <- mean(x, na.rm = TRUE)+(sd(x, na.rm = TRUE)*0.25)
  sd1n <- mean(x, na.rm = TRUE)-(sd(x, na.rm = TRUE)*0.25)
  sd2p <- mean(x, na.rm = TRUE)+(sd(x, na.rm = TRUE)*1)
  sd2n <- mean(x, na.rm = TRUE)-(sd(x, na.rm = TRUE)*1)
  ifelse(x > sd2p, 4,
         ifelse(x > sd1p & x < sd2p, 3,
                ifelse(x > sd1n & x < sd1p, 2,
                       ifelse(x > sd2n & x < sd1n, 1, 0))))
}

#### TidyCensus data wrangling and modification ########
tidymut <- tidyadd %>%
  mutate(his_per = round(his_popE/tot_popE, digits = 2)*100,
         bl_per = round(bl_popE/tot_popE, digits = 2)*100,
         wh_per = round(wh_popE/tot_popE, digits = 2)*100,
         as_per = round(as_popE/tot_popE, digits = 2)*100,
         inc_qc = sdcut(med_incE),
         inc_gr = ifelse(inc_qc == 0, "Very Below", 
                         ifelse(inc_qc == 1, "Somewhat Below",
                                ifelse(inc_qc == 2, "About Average",
                                       ifelse(inc_qc == 3, "Somewhat Above",
                                              ifelse(inc_qc == 4, "Very Above", "NA"))))),
         maxdemo = pmax(his_per, bl_per, wh_per, as_per),
         demomaj = ifelse(his_per == maxdemo & maxdemo >= 45, "Hispanic",
                          ifelse(bl_per == maxdemo & maxdemo >= 45, "Black",
                                 ifelse(wh_per == maxdemo & maxdemo >= 45, "White",
                                        ifelse(as_per == maxdemo & maxdemo >= 45, "Asian", "Multi")))),
         es3t4_per = round(es_3t4E/pop_3t4E, digits = 2)*100,
         bp = round(pop_bpE/tot_popE, digits = 2)*100,
         cpr = round(bp_u18E/pop_u18E, digits = 2)*100,
         cpr_u5 = round(bp_u5E/pop_u5E, digits = 2)*100,
         rohh_per = round(rohhE/thhE, digits = 2)*100,
         oohh_per = round(oohhE/thhE, digits = 2)*100,
         pubas_u18per = round(pubas_u18E/thhE, digits = 2)*100,
         pubas_per = round(pubas_totE/thhE, digits = 2)*100,
         births_u100 = birmar_u100E+birunm_u100E,
         births_bp_per = round(births_u100/wraE, digits = 2)*100,
         births_per = round(birthsE/wraE, digits = 2)*100,
         hs_per = hs_degE/pop_degE,
         ba_per = ba_degE/pop_degE
         ) %>%
  select(-ends_with("M"), -maxdemo, -inc_qc, -(birmar_u100E:birunm_u100E), pop_degE)

medinc_vb <- tidymut %>%
  filter(inc_gr == "Very Below")

medinc_sb <- tidymut %>%
  filter(inc_gr == "Somewhat Below")

medinc_aa <- tidymut %>%
  filter(inc_gr == "About Average")

medinc_sa <- tidymut %>%
  filter(inc_gr == "Somewhat Above")

medinc_va <- tidymut %>%
  filter(inc_gr == "Very Above")

medinc_na <- tidymut %>%
  filter(is.na(inc_gr))
```

```{r, fig.width=9, echo=FALSE, message=FALSE, warning=FALSE}
#### Mapbox Styling link and leaflet ###################
cpal_style <- "https://api.mapbox.com/styles/v1/owencpal/ckecb71jp22ct19qc1id28jku/tiles/256/{z}/{x}/{y}@2x?access_token=pk.eyJ1Ijoib3dlbmNwYWwiLCJhIjoiY2tlYnR3emdxMGNhZzMwb2EzZWR4ajloNCJ9.P7Mujz8F3Rssq5-Q6dcvMw"

map_attr <- "<a href='https://www.mapbox.com/map-feedback/'>Mapbox</a> Basemap <a href='https://childpovertyactionlab.org/'>Child Poverty Action Lab</a>"

factdem <- colorFactor(CPAL.colors, 
                       levels = c("Asian", "Black", "Hispanic", NA, "Multi", "White"))

popup_vb <- paste0("<b>", "GEOID: ", "</b>", medinc_vb$GEOID, "<br>",
               "<b>", "Median Income: ", "</b>", "$", medinc_vb$med_incE, "<br>",
               "<b>", "Total Population: ", "</b>", medinc_vb$tot_popE, "<br>",
               "<b>", "Asian (%): ", "</b>", medinc_vb$as_per, "%", "<br>",
               "<b>", "Black (%): ", "</b>", medinc_vb$bl_per, "%", "<br>",
               "<b>", "Hispanic (%) : ", "</b>", medinc_vb$his_per, "%", "<br>",
               "<b>", "White (%): ", "</b>", medinc_vb$wh_per, "%", "<br>",
               "<b>", "Child Poverty Rate (u5): ", "</b>", medinc_vb$cpr_u5, "%", "<br>",
               "<b>", "Women Births u100% FPL: ", "</b>", medinc_vb$births_bp_per, "%", "<br>",
               "<b>", "Households w/u18 Receiving SNAP/Food Stamps: ", "</b>", medinc_vb$births_bp_per, "%")

popup_sb <- paste0("<b>", "GEOID: ", "</b>", medinc_sb$GEOID, "<br>",
                "<b>", "Median Income: ", "</b>", "$", medinc_sb$med_incE, "<br>",
                "<b>", "Total Population: ", "</b>", medinc_sb$tot_popE, "<br>",
                "<b>", "Asian (%): ", "</b>", medinc_sb$as_per, "%", "<br>",
                "<b>", "Black (%): ", "</b>", medinc_sb$bl_per, "%", "<br>",
                "<b>", "Hispanic (%) : ", "</b>", medinc_sb$his_per, "%", "<br>",
                "<b>", "White (%): ", "</b>", medinc_sb$wh_per, "%", "<br>",
                "<b>", "Child Poverty Rate (u5): ", "</b>", medinc_vb$cpr_u5, "%", "<br>",
                "<b>", "Women Births u100% FPL: ", "</b>", medinc_vb$births_bp_per, "%", "<br>",
                "<b>", "Households w/u18 Receiving SNAP/Food Stamps: ", "</b>", medinc_vb$births_bp_per, "%")

popup_aa <- paste0("<b>", "GEOID: ", "</b>", medinc_aa$GEOID, "<br>",
                "<b>", "Median Income: ", "</b>", "$", medinc_aa$med_incE, "<br>",
                "<b>", "Total Population: ", "</b>", medinc_aa$tot_popE, "<br>",
                "<b>", "Asian (%): ", "</b>", medinc_aa$as_per, "%", "<br>",
                "<b>", "Black (%): ", "</b>", medinc_aa$bl_per, "%", "<br>",
                "<b>", "Hispanic (%) : ", "</b>", medinc_aa$his_per, "%", "<br>",
                "<b>", "White (%): ", "</b>", medinc_aa$wh_per, "%", "<br>",
                "<b>", "Child Poverty Rate (u5): ", "</b>", medinc_vb$cpr_u5, "%", "<br>",
                "<b>", "Women Births u100% FPL: ", "</b>", medinc_vb$births_bp_per, "%", "<br>",
                "<b>", "Households w/u18 Receiving SNAP/Food Stamps: ", "</b>", medinc_vb$births_bp_per, "%")

popup_sa <- paste0("<b>", "GEOID: ", "</b>", medinc_sa$GEOID, "<br>",
                "<b>", "Median Income: ", "</b>", "$", medinc_sa$med_incE, "<br>",
                "<b>", "Total Population: ", "</b>", medinc_sa$tot_popE, "<br>",
                "<b>", "Asian (%): ", "</b>", medinc_sa$as_per, "%", "<br>",
                "<b>", "Black (%): ", "</b>", medinc_sa$bl_per, "%", "<br>",
                "<b>", "Hispanic (%) : ", "</b>", medinc_sa$his_per, "%", "<br>",
                "<b>", "White (%): ", "</b>", medinc_sa$wh_per, "%", "<br>",
                "<b>", "Child Poverty Rate (u5): ", "</b>", medinc_vb$cpr_u5, "%", "<br>",
                "<b>", "Women Births u100% FPL: ", "</b>", medinc_vb$births_bp_per, "%", "<br>",
                "<b>", "Households w/u18 Receiving SNAP/Food Stamps: ", "</b>", medinc_vb$births_bp_per, "%")

popup_va <- paste0("<b>", "GEOID: ", "</b>", medinc_va$GEOID, "<br>",
                "<b>", "Median Income: ", "</b>", "$", medinc_va$med_incE, "<br>",
                "<b>", "Total Population: ", "</b>", medinc_va$tot_popE, "<br>",
                "<b>", "Asian (%): ", "</b>", medinc_va$as_per, "%", "<br>",
                "<b>", "Black (%): ", "</b>", medinc_va$bl_per, "%", "<br>",
                "<b>", "Hispanic (%) : ", "</b>", medinc_va$his_per, "%", "<br>",
                "<b>", "White (%): ", "</b>", medinc_va$wh_per, "%", "<br>",
                "<b>", "Child Poverty Rate (u5): ", "</b>", medinc_vb$cpr_u5, "%", "<br>",
                "<b>", "Women Births u100% FPL: ", "</b>", medinc_vb$births_bp_per, "%", "<br>",
                "<b>", "Households w/u18 Receiving SNAP/Food Stamps: ", "</b>", medinc_vb$births_bp_per, "%")

#### Leaflet Map ###################
leaflet() %>%
  setView(lng = -96.7970, lat = 32.7767, zoom = 10) %>%
  addTiles(urlTemplate = cpal_style, attribution = map_attr) %>%
  addPolygons(data = medinc_vb,
              fillColor = ~factdem(demomaj),
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              group = "Very Below",
              popup = ~popup_vb) %>%
  addPolygons(data = medinc_sb,
              fillColor = ~factdem(demomaj),
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              group = "Somewhat Below",
              popup = ~popup_sb) %>%
  addPolygons(data = medinc_aa,
              fillColor = ~factdem(demomaj),
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              group = "About Average",
              popup = ~popup_aa) %>%
  addPolygons(data = medinc_sa,
              fillColor = ~factdem(demomaj),
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              group = "Somewhat Above",
              popup = ~popup_sa) %>%
  addPolygons(data = medinc_va,
              fillColor = ~factdem(demomaj),
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              popup = ~popup_va,
              group = "Very Above") %>%
  addPolygons(data = medinc_na,
              fillColor = "#b4b4b4",
              weight = 1,
              opacity = 0.6,
              color = "#FFFFFF",
              fillOpacity = 0.6,
              highlight = highlightOptions(
                weight = 4,
                color = "#FFFFFF",
                fillOpacity = 0.4,
                bringToFront = TRUE),
              group = "NA",
              popup = "NA") %>%
  addLegend(data = tidymut,
            pal = factdem, 
            values = ~demomaj, 
            opacity = 0.7, 
            title = "Majority Demographic",
            position = "bottomright") %>%
  addLayersControl(
    overlayGroups = c("Very Below", "Somewhat Below", "About Average", "Somewhat Above", "Very Above", "NA"),
    options = layersControlOptions(collapsed = FALSE),
    position = "bottomleft"
  ) %>%
  htmlwidgets::onRender("
        function() {
            $('.leaflet-control-layers-overlays').prepend('<label style=\"text-align:center\">Median Income</label>');
        }
    ")
```

## Using the Map {-}

The toggles on the top right of the map have grouped all census tracts based on how the median income in a census tract differs from the mean income in Dallas County.
I.E. 

* __Very Below:__ Contains all tracts greater than 1 standard deviation below the mean.

* __somewhat Below:__ Contains all tracts between -1 and -0.25 standard deviations below the mean.

* __About Average:__ Contains all tracts between -0.25 and 0.25 standard deviations from the mean.

* __Somewhat Above:__ Contains all tracts between 0.25 and 1 standard deviations above the mean.

* __Very Above:__ Contains all tracts greater than 1 standard deviation above the mean.

The intention behind the toggles is to visually identify neighborhoods which may have similar demographic and racial/ethnic makeups for on the ground exploration.

The toggles can be generated based on other variables such as Child Poverty Rate upon request if a different view is wanted.

Clicking/Hovering on an individual tract will lower the opacity allowing you to view the streets an intersections of the tract as well as providing a small popup containing some basic information about the census tract including demographic, and economic data.

# Key Birth to 3 Metrics {-}

The table below contains a subsection of data pulled for this data request allowing a quick visual reference for some key data of interest.
Table can be sorted in ascending or descending order based on any variable using the arrows.
If you would like to filter the table based on a subset of any values click on the box below the variable name and a range bar will appear allowing you to make those selections.

```{r, fig.height=12, fig.width=9, echo=FALSE, message=FALSE, warning=FALSE}
namescol <- c("GEOID", "Total Population", "Hispanic (%)", "Black (%)", "White (%)", "Asian (%)", "Majority Demographic", "Median Income", "Child Poverty Rate (u18)", "Child Poverty Rate (u5)", "Women Birth 100% FPL", "Enrolled Pre-School (%)", "Total Child Care Centers", "Average Household Size", "Living with Grandparents u6", "Households u18 Food Stamps/SNAP (%)")
tidymut %>%
  as.data.frame(.) %>%
  select(GEOID, tot_popE, his_per:as_per, demomaj, med_incE, cpr, cpr_u5, births_bp_per, es3t4_per, tot_ccc, avg_hhsE, gran_u6E, pubas_u18per) %>%
  DT::datatable(., class = 'cell-border stripe', colnames = namescol, rownames = FALSE, options = list(
  pageLength=15, scrollX='400px'), filter = 'top')
```

## Data Definitions {-}
* __GEOID:__ Unique identifier of census tract.

* __Total Population:__ Total population count.

* __Hispanic (%):__ Percent Hispanic population

* __Black (%):__ Percent Black population

* __White (%):__ Percent White Non-Hispanic population

* __Asian (%):__ Percent Asian population

* __Majority Demographic:__ Demographic that is at least 45% of the population of a tract.

* __Median Income:__ Median income of census tract

* __Child Poverty Rate (u18):__ Percent children under 18 living below poverty.

* __Child Poverty Rate (u5):__ Percent children under 5 living below poverty.

* __Women Birth 100% FPL:__ Percent of women who have given birth within the last 12 months who are below 200% of the federal poverty line.

* __Enrolled Pre-School:__ Percent of children between 3 and 4 enrolled in school.

* __Child Households Public Assistance (%):__ Percent of children under 18 who live in households receiving public assistance (includes SNAP, WIC, etc.)

* __Total Child Care Centers:__ Total number of child care centers per Texas Department of State Health Services

* __Average Household Size:__ Average household size

* __Living with Grandparents u6:__ Total number of children under 6 living in grandparents household.

* __Households w/u18 Receiving Food Stamps/SNAP:__ Percent of households with children under 18 receiving food stamps/SNAP.